
# Dose events {#dose-events-chapter}

```{r, include = FALSE}
source("setup.R")
```

mrgsolve can implement both bolus and infusion doses. Ordinarily, this is 
accomplished through an input data set (@sec-data-set-chapter and 
@sec-event-chapter). Doses can also be implemented from within the model 
itself (@sec-plugin-evtools). 


## Bolus dose

A bolus dose is implemented by adding the dose amount to the indicated 
compartment. To implement a bolus dose, use event ID `1`. 

```{r}
#| message: false
mod <- modlib("pk1cmt", end = 72, delta = 0.1)

dose <- ev(time = 10, evid = 1, amt = 100, cmt = 1)

mrgsim(mod, dose) %>% plot()
  
```

In this example, to give the dose at 10 hours, the system is advanced to the 
time of the dose record and the 100 unit dose amount is added to the first 
compartment. Before the problem continues, the ODE solver is reset, creating 
a discontinuity in the simulation. Specifically, the solver restarts with a 
new initial value (the amount already in `EV` at the time of the dose plus
the added dose amount). 

::: {.callout-note}
# Event object construction
In the previous example, we specified each argument in the `ev()` 
constructor. All of the arguments were specified at their default values 
(e.g., `time = 0`, `evid = 1`) for clarity. The same event object can be 
created with 

```{r}
#| eval: false
dose <- ev(amt = 100)
```

See @sec-event-chapter for more information.
:::


## Intermittent infusion dose

To implement a single, intermittent infusion dose, specify the `rate` in the 
data set or event object. We can convert the previous bolus dose as an 
infusion like this

```{r}
dose <- ev(time = 10, evid = 1, amt = 100, rate = 10, cmt = 1)

mrgsim(mod, dose) %>% plot()
```

Because there was a 100 mg dose infused at a rate of 10 mg/hr, the duration 
of the infusion is 10 hours. In this case, mrgsolve integrates up to the time 
of the infusion dose and starts a zero-order infusion at the specified rate. 
At the time the dose is started, mrgsolve schedules another event to turn the 
infusion off. Discontinuities (ODE solver resets) are introduced in the
simulation at the times when infusions start or end.

While infusions are sometimes thought of as extra-vascular administration 
(like an IV infusion), these doses can be administered to any compartment in 
the model. 

This is an intermittent dose because it has a definite start and end time.

### Modeled infusion rate and duration

In certain cases, we'd like to model the infusion rate or the infusion duration.
In that case, we cannot specify `rate` in the data set prior to the simulation;
it must be set in the model itself. 

To do this, enter the infusion rate or duration in the `$PK` block. We'll 
expand the example to model the duration of an infusion over 20 hours. We 
write the model with `D_DEPOT = 20;` to indicate that any infusion in to the 
`DEPOT` compartment should have a duration of 20 hours. 


```{r}
#| message: false

code <- '
$PARAM CL = 1, V = 20, KA = 1
$PKMODEL cmt = "DEPOT,CENT", depot = TRUE
$PK
D_DEPOT = 20;
'

mod <- mcode("dose-events-1", code, end = 72, delt = 0.1)
```

To model the infusion rate, use `R_DEPOT = 5;` to get a duration of 20 hours 
for 100 mg dose.


::: {.callout-note}
`R_DEPOT` and `D_DEPOT` are the original variables to set for modeling infusion
rate and duration, respectively. NONMEM uses `Rn` and `Dn` instead, where 
`n` is the compartment number (i.e. `R1` and `D1`). To use this syntax instead,
you need to invoke the `nm-vars` plugin (@sec-plugin-nm-vars).
:::

When modeling the infusion duration or rate, you must set `RATE` to `-1` 
(modeled infusion rate) or `-2` (modeled infusion duration) in your input 
data set or event object. In our case we modify the `rate` item in the event 
object to `-2` because we are modeling the infusion duration.

```{r}
dose_d <- mutate(dose, rate = -2)

mrgsim(mod, dose_d) %>% plot()
```

mrgsolve will warn you if you have not provided a `RATE` data item that is 
consistent with your model. So, we would get warned if we forgot to set 
`RATE = -2`.

```{r}
out <- mrgsim(mod, dose)
```

Note that it is not unreasonable to request either / both bolus or infusion 
doses into a compartment with a modeled infusion duration or rate. To enable 
this behavior, you can either write logic into `$PK` that detects the dose
type and set either `R_CMT` or `D_CMT` to a positive number for an infusion
or 0 for bolus; this will silence the warning from mrgsolve. A more global 
option is to set the `@!check_modeled_infusions` option on the `$PK` (or 
`$MAIN`) block. 


