```{r,include = FALSE}
source("setup.R")
```

# Plugins {#sec-plugins}

## autodec {#sec-plugin-autodec}

Available as of mrgsolve version 1.0.0.

When this plugin is invoked, mrgsolve will search your model code for assignments
and automatically declare them as `double` precision numbers. The following 
blocks are searched

- `$PREAMBLE`
- `$MAIN` (or `$PK`)
- `$ODE` (or `$DES`)
- `$TABLE` (or `$ERROR`)
- `$PRED`

For example, the following code requires that `CL` gets assigned a type

```{c, eval = FALSE}
$PARAM WT = 70, TVCL = 1.2

$PK
double CL = TVCL * pow(WT/70, 0.75);
```

This is the default mrgsolve behavior and has been since the beginning.

The `autodec` plugin lets you write the following

```{c, eval = FALSE}
$PLUGIN autodec 

$PARAM WT = 70, TVCL = 1.2

$PK
CL = TVCL * pow(WT/70, 0.75);
```

mrgsolve will find `CL = ...` and understand that this is a user initiated 
variable and will declare it as `double` for you. Don't worry about `WT = 70` 
in `$PARAM`; mrgsolve should already know about that won't try to declare it. 

When you are using the `autodec` plugin, you can still declare variables as 
`double` or `int` or `bool`. mrgsolve already finds those variables and will 
understand to leave those declarations alone. Note that it may still very 
convenient to declare using the `capture` type those variables that you 
want captured into the output

```{c, eval = FALSE}
$PLUGIN autodec

$ERROR
capture Y = IPRED * exp(EPS(1));
```

The `capture` typedef makes `Y` a `double`; we didn't need to declare it with 
`autodec` in play, but decided to declare with `capture` so that it is copied
into the simulated output.


The `autodec` plugin is intended for more straightforward models where most / 
all variables are real valued. Because mrgsolve can handle any valid C++ code
in these blocks, there is a possibility that the code could get much more 
complicated, including custom classes and methods. In this case, we recommend
to bypass this feature and take control of declaring variables as you would 
in the default mode. 

In case mrgsolve does try to declare (as `double`) a variable that shouldn't 
be handled that way, you can note this name in an environment variable inside
your model called `MRGSOLVE_AUTODEC_SKIP`

```{c, eval = FALSE}
$ENV MRGSOLVE_AUTODEC_SKIP = c("my_variable_1")
```

This can be a vector of variable names to NOT declare when `autodec` is invoked. 

## nm-vars {#sec-plugin-nm-vars}

Available as of mrgsolve version 1.0.0.

The `nm-vars` plugin provides a more NONMEM-like set of macros to use when 
coding your compartmental model. Only a small subset of the NONMEM model syntax
is replicated here. 


__F, R, D, ALAG__

- To set bioavailability for the nth compartment, use `Fn`
- To set the infusion rate for the nth compartment, use `Rn`
- To set the infusion duration for the nth compartment, use `Dn`
- To set the lag time for the nth compartment, use `ALAGn`

For example

```{c, eval = FALSE}
$CMT GUT CENT GUT2

$PK
F1 = 0.87;    // equivalent to F_GUT  = 0.87;
R2 = 2.25;    // equivalent to R_CENT = 2.25;
ALAG3 = 0.25; // equivalent to ALAG_GUT2 = 0.25; 
```

__A, A_0, DADT__

- To refer to the amount in the nth compartment, use `A(n)`
- To refer to the initial amount in the nth compartment, use `A_0(n)`
- To refer to the differential equation for the nth compartment, use `DADT(n)`

For example

```{c, eval = FALSE}
$CMT CMT1 CMT2

$PK
A_0(2) = 50;
  
$DES
DADT(1) = -KA * A(1);
DADT(2) =  KA * A(1) - KE * A(2); 
```

__Math__

Starting with version 1.0.1, macros are provided for several math functions

- `EXP(a)`  gets mapped to `exp(a)`
- `LOG(a)`  gets mapped to `log(a)`
- `SQRT(a)` gets mapped to `sqrt(a)`

These are purely for convenience, so that upper-case versions from NMTRAN don't
require conversion to lower-case; this happens automatically via the C++ 
preprocessor. 

__Other syntax__

- Using `THETA(n)` in model code will resolve to `THETAn`; this feature is 
  always available, even when `nm-vars` hasn't been invoked; we mention it here
  since it is a fundamental piece of the NONMEM syntax that mrgsolve has 
  internalized 
- Use `T` in `$DES` to refer to the current time in the odesolver rather
  than `SOLVERTIME`
  
__Reserved words with nm-vars is invoked__

There are some additional reserved words when the `nm-vars` plugin is invoked

- `A`
- `A_0`
- `DADT`
- `T`

It is an error to use one of these symbols as the name of a parameter or 
compartment or to try to declare them as variables. 

__mrgsolve syntax that is still required__

There are a lot of differences remaining between mrgsolve and NONMEM syntax. We
mention a few here to make the point

- mrgsolve continues to require `pow(base, exponent)` rather than `base**exponent`
- mrgsolve continues to require a semi-colon at the end of each statement (this
  is a C++ requirement)
- mrgsolve continues to require that user-defined variables are declared with 
  a type, except when the `autodec` plugin (@sec-plugin-autodec) is invoked
  
__An example__

There is an example of this syntax (along with `autodec` features) in the 
internal model library

```{r, eval = FALSE}
mod <- modlib("nm-like")
see(mod)
```

```{r, echo = FALSE}
see(modlib("nm-like", compile = FALSE))
```


## tad {#sec-plugin-tad}

__Purpose__
Advanced calculation time after dose within your model. We call this "advanced"
because it lets you track doses in multiple compartments. See the note below 
about a simpler way to calculate time after dose that should work fine if doses
are only in a single compartment. This functionality is provided by mrgsolve. 

__Usage__

First, tell `mrgsolve` that you want to use the `tad` plugin

```{c, eval = FALSE}
$PLUGIN tad
```

The create `tadose` objects, one for each compartment where you want to track
time after dose. One approach is to do this in `[ global ]`

```{c, eval = FALSE}
[plugin] tad

[ global ] 
mrg::tadose tad_cmt_1(1); 
mrg::tadose tad_cmt_2(2);
```

Notice that we pass the compartment number that we want to track in each 
case and also that we refer to the `mrg::` namespace for the `tadose` class. 

The `tadose` objects contain the following (public) members

- `cmt` the compartment to track
- `told` the time of last dose; defaults to `-1e9`
- `had_dose` indicates if a dose has already been given for the current 
  individual
- `tad(self)` the function to call to calculate time after dose
  - the `self` object (@sec-self-object) must be passed as the only
    argument
  - when the member function is called prior to the first administered dose, a 
    value of `-1.0` is returned
- `reset()` resets the state of the object; be sure to reset prior to simulating
  a new individual

As an example, you can call the `reset()` method on one of the `tadose` objects

```{c, eval = FALSE}
tad_cmt_1.reset();
```


You can find the source code for this object 
[here](https://github.com/metrumresearchgroup/mrgsolve/blob/master/inst/base/mrgsolve-tad.h).

A working example model that tracks doses in compartments `1` and `2` is
provided here

```{c, eval = FALSE}
[plugin] tad

[ global ] 
mrg::tadose tad_cmt_1(1); 
mrg::tadose tad_cmt_2(2);

[ pkmodel ] cmt = "GUT,CENT", depot = TRUE

[ param ] CL = 1, V = 20, KA = 1

[ main ] 
capture tad1 = tad_cmt_1.tad(self); 
capture tad2 = tad_cmt_2.tad(self);
```

__Static approach__

Another approach would be to make these static in `[ main ]` but this approach
would only work if you only use these in `[ main ]`; the `[ global ]` approach 
is preferable since then you can access the object in any block (function). 


### Note
Note there is a simpler way to calculate time after dose when only dosing into
a single compartment

```{c, eval = FALSE}
[ main ]
double tad = self.tad();
```

The `self` object (@sec-self.tad) contains a `tad()` member which 
will track time after dose. Note that this needs to be called every record. 


## evtools {#sec-plugin-evtools}

__Purpose__

The `evtools` plugin is a set of functions and classes you can use to implement
dosing regimens from _inside_ your model. It first became available in mrgsolve
1.4.1. The most common use for this plugin is when you want to implement dynamic
dosing simulations where the dose amount or the dosing interval is able to
change based on how the system has advanced up to a certain point. For example,
you might have a PKPD model for an oncology drug that includes a PK model for
the drug as well as a dynamic model for platelets where a decline in platelets
is driven by the drug concentration. In this case you might monitor platelets at
different clinical visits and reduce the or hold dose or increase the dosing
interval in response to Grade 3 or Grade 4 thrombocytopenia.

__Usage__

Like all other plugins, you must invoke `evtools` in the `$PLUGIN` block

```{c, eval = FALSE}
$PLUGIN evtools
```

### Namespace

All the functionality made available by the `evtools` plugin is located in a 
namespace called `evt`. So you will need to prefix all functions and classes
with `evt::`. For example, you can read about a function called `bolus` below; 
when you call that function, you need to refer to `evt::bolus`, locating that 
function in the `evt` namespace. 

### Event object type

@sec-mtime introduces a C++ event object called `mrg::evdata`. The `evt` 
namespace provides an easier to remember `typedef` for that object called
`evt::ev`. So if a function returns an event object, you can use `evt::ev`
for that type. For example

```{c}
#| eval: false
evt::ev dose = evt::bolus(100, 1);
```

is equivalent to 

```{c}
#| eval: false
mrg::evdata dose evt::bolus(100, 1);
```

### Simple administration of single doses now

The `evt` namespace allows you to easily administer single bolus or infusion
doses. The functions are

- `void evt::bolus(self, <amt>, <cmt>)` where 
  - `self` is the the self object, described in @sec-self-object
  - `<amt>` is the dose amount (type `double`)
  - `<cmt>` is the dosing compartment (type `double`)

- `void evt::infuse(self, <amt>, <cmt>, <rate>)` where 
  - `self` is the the self object, described in @sec-self-object
  - `<amt>` is the dose amount (type `double`)
  - `<cmt>` is the dosing compartment (type `double`)
  - `<rate>` is the infusion rate (type `double`)

Note that `self` is passed as the first argument, there is no return value, 
and there is no `time` specified for the dose. All doses invoked this way are 
given `now`, as-is; so you should only call these functions when the model 
code has decided it is time to administer a dose. 

_Important_: Because doses are given `now`, these functions should almost be 
called in `$TABLE` (i.e., `$ERROR`).

### Customized doses, potentially given later

The `evt` namespace also provides variants of these functions which return 
the event object to you so you can modify some of the attributes, including 
potentially scheduling the dose in the future, prior to sending the object
back to mrgsolve. These functions are 

- `evt::ev evt::bolus(<amt>, <cmt>)` where 
  - `<amt>` is the dose amount (type `double`)
  - `<cmt>` is the dosing compartment (type `double`)

- `evt::ev evt::infuse(<amt>, <cmt>, <rate>)` where 
  - `<amt>` is the dose amount (type `double`)
  - `<cmt>` is the dosing compartment (type `double`)
  - `<rate>` is the infusion rate (type `double`)

Note that we _don't_ pass in the `self` object here; just the dose amount, 
compartment, and rate for infusions. These functions also return and event 
object (type `evt::ev`) that you can work with. See @sec-modeled-doses
for documentation of those attributes. 

### API for customizing doses

While @sec-modeled-doses shows you some low-level ways to customize the event 
object, the `evt` namespace provides some API for making these changes. 

__`evt::retime`__

The `evt::retime` function can be used to set the time attribute. 

```{c}
#| eval: false
evt::ev dose = evt::bolus(100, 1);
evt::retime(dose, 24);
self.push(dose);
```

**Arguments**: 

- an event object (`evt::ev`)
- the new dose time (`<double>`)

**Return**: `void` (or nothing)

When doses are retimed this way, the `now` attribute is forced to be `false`.

__`evt::now`__

Use `evt::now` to set the `now` attribute to `true`

```{c}
#| eval: false
evt::now(dose);
```

**Argument**: 

- an event object (`evt::ev`)

**Return**: `void` (nothing)

__`evt::push`__

The `evt` namespace includes a `push` function to send an event object back 
to mrgsolve. For example

```{c}
#| eval: false
evt::ev dose = evt::bolus(100, 1);
evt::retime(dose, 24);
evt::push(self, dose);
```

This function will continue to be available in the `evt` namespace. But note 
that `self` has a `push()` method as of mrgsolve 1.4.1 to do the same thing

__`evt::near()`__

Use this function to test for equality between floating put numbers. For 
example, to test if `TIME` is (about) equal to `24.5`, you can call

```{c}
#| eval: false
if(evt::near(TIME, 24.5)) { 
  // do something  
}
```

This function is similar to the `dplyr::near()` function.

**Arguments**: 

- a number to test (`double`)
- another number to test (`double`)
- optional argument `<eps>`, which is the tolerance for establishing equality
  between the two test numbers; `<eps>` defaults to `1e-8` 

**Return**: `bool`

### Class to implement a dosing regimen

The `evtools` namespace also includes a class for implementing "automatic" 
dosing in a regimen. The documentation presented here will be limited to 
a brief discussion of the constructor and member functions for this class. 
More is written in @sec-mtime about how you can use this class effectively. 

- The constructor `evt::regimen::regimen()` does not take any arguments, but 
  it dose call the `reset()` method. 
- `void init(self)` initializes the object; the argument is the `self` object
  (see @sec-self-object)
- `void reset()` resets the object to sensible defaults
  - dose compartment is set to 1
  - dose amount is set to 0
  - infusion rate is set to 0
  - dosing interval is set to 1e9
  - dosing duration is set to 1e9
  - other internal configuration

A series of setter functions let you set different attributes for the dosing 
regimen. All of the following functions return `void`. In the examples below, 
`object` refers to an object with class `evt::regimen`. 

- `object.amt(<double>)` sets the dose amount
- `object.cmt(<int>)` sets the dosing compartment number
- `object.rate(<double>)` sets the infusion rate
- `object.ii(<double>)` sets the dosing interval
- `object.until(<double>)` sets the time of the last dose

Similarly, there are a set of getter functions to return these data members

- `double object.amt()` returns the dose amount
- `int object.cmt()` returns the dosing compartment number
- `double object.rate()` returns the infusion rate
- `double object.ii()` returns the dosing interval
- `double object.until()` return the time of the last dose

To start the dose regimen, call 

```{c}
#| eval: false
object.execute()
```

This should almost always be called in `$TABLE` (i.e., `$ERROR`).


To force the simulation to stop at the time of the next dose with `EVID` set 
to 3333, use the `flagnext()` member function

```{c}
#| eval: false
object.flagnext();
```

This is usually set once at the start of the problem, either in `$PREAMBLE` or 
in `$MAIN` when `NEWIND <= 1`.


## CXX11  {#sec-plugin-cxx11}

__Purpose__

Compile your model file with `C++11` standard.

__Usage__ 

```{c, eval = FALSE}
$PLUGIN CXX11
```

## Rcpp {#sec-plugin-rcpp}

__Purpose__

Link to `Rcpp` headers into your model.

__Usage__ 

```{c, eval = FALSE}
$PLUGIN Rcpp
```

Note that once your model is linked to `Rcpp`, you can start using that 
functionality immediately (without including `Rcpp.h`).

A very useful feature provided by `Rcpp` is that it exposes all of the `dpqr`
functions that you normally use in R (e.g. `rnorm()` or `runif()`). So, if you
want to simulate a number from Uniform (0,1) you can write

```{c, eval = FALSE}
$PLUGIN Rcpp

$TABLE
double uni = R::runif(0,1);
```

Note that the arguments are the same as the R version (`?runif`) **except** 
there is no `n` argument; you always only get one draw.

Information about `Rcpp` can be found here: https://github.com/RcppCore/Rcpp

## mrgx {#sec-plugin-mrgx}

Compile in extra C++ / Rcpp functions that can be helpful to you 
for more advanced model coding. The `mrgx` plugin is dependent on the
`Rcpp` plugin. 

The functions provided by `mrgx` are in a namespace of the same name, 
so to invoke these functions, you always prepend `mrgx::`. 


### Get the model environment

Note that your model object (`mod`) contains an R environment. For 
example

```{r}
mrgsolve::house()@envir
```

The objects in this environment are created by a block called `$ENV`
in your model code (see @sec-block-env);


To access this environment in your model, call

```{c, eval = FALSE}
Rcpp::Environment env = mrgx::get_envir(self);
```

## Extract an object from the model environment

When you have an object created in `$ENV`

```{c, eval = FALSE}
[ env ] 
rand <- rnorm(100)
```

You can extract this object with

```{c, eval = FALSE}
[ preamble ]
Rcpp::NumericVector draw = mrgx::get("rand", self);
```



## RcppArmadillo

__Purpose__

Link to `RcppArmadillo` headers into your model. 

__Usage__ 

```{c, eval = FALSE}
$PLUGIN RcppArmadillo
```

Information about `armadillo` can be found here: http://arma.sourceforge.net/
Information about `RcppArmadillo` can be found here: 
https://github.com/RcppCore/RcppArmadillo

## BH

__Purpose__

Link to `boost` headers into your model. 

__Usage__ 

```{c, eval = FALSE}
$PLUGIN BH
```

Note that once your model is linked to `BH` (`boost`), you will be able to 
include the `boost` header file that you need. You have to include the header
file that contains the `boost` function you want to use.

Information about `boost` can be found here: https://boost.org. Information 
about `BH` can be found here: https://github.com/eddelbuettel/bh
